---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

```{r include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

# sev - The SummarizedExperiment Viewer

## Overview
The sev package is a utility package that relies completely on the bioconductor packages DEP and iSEE/iSEEu for the analysis, adding only convenience functions, imputation methods and additional iSEE containers.

## Installation
Open up R and run the following code:

```{r eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
    
BiocManager::install("iSEE")
BiocManager::install("iSEEu")

#All other dependencies are installed automatically with sev. 
devtools::install_github("adhutz/sev")
```

## Data import
Data can be imported directly from MaxQuant derived proteinGroups.txt or similar files to create a summarized experiment. 

### Possibility 1: Read file in and "guess" sample names
lfq columns must begin with "lfq_intensity" and end with **sep** followed by replicate number.

Example: "lfq_intensity_ctrl_rep_1"

```{r import-data}
library(sev)
library(DEP)
library(dplyr)
library(clusterProfiler)
library(dplyr)
library(fdrtool)
library(ggplot2)
library(ggrepel)
library(HybridMTest)
library(janitor)
library(SingleCellExperiment)
library(splitstackshape)
library(stats)
library(tidyr)
library(magrittr)

se <- sev::se_read_in("../data/example_1_proteinGroups.txt", sep = "_rep_")

se 
```

### Possibility 2: Read exerimental design from file
For this, the design file must contain the columns "sample", "label", "condition", and "replicate". The column sample must contain the name of the column that stores the quantitative data (e.g. lfq values).

```{r import-data2}
design <- read.csv(file = "../data/example_1_design.txt")

design

se <- sev::se_read_in("../data/example_1_proteinGroups.txt", experimental_design = design)

se
```

## Filter
MS data can contain a lot of missing values. Ultimately, a part can be estimated and imputed based on other values. Before that, it is necessary to filter out rows with too many missing values. The user can specify the fraction of missing values that is tolerated. The condition must be fulfilled in each group, at least one group or across all samples, depending on the *filter_mode*.

```{r filter}
# Filter out samples with too many missing values
se_filtered <- sev::filter_perseus(se, perc_na = 0.33, filter_mode = "each_group")
```

## Imputation
Now that rows with too many missing values are excluded, we can replace NAs with values. The *DEP::impute()* function already provides a variety of options. The *sev* package builds on that, but saves the raw data including NAs in another assay instead of overwriting it. In addition, a binary assay with 1 for imputed values and 0 for measured values is created to easily visualize missing data via iSEE afterwards. 

### Possibility 1: Use DEP standard functions
```{r impute DEP}
se_imp <- sev::impute_DEP(se_filtered, fun="QRILC")
```

### Possibility 2: Use mixed imputation for **MAR** vs. **MNAR**
It is possible to distinguish different types of "missingness", termed either **missing at random (MAR)** or **missing not at random (MNAR)**. The *DEP* package already allows to select different imputation methods depending on if rows contain **MAR** or **MNAR** values via the *mixed* argument. For this, the column *randna* is necessary in the rowData that specifies the type of missingness. This column can now be added via the **sev::add_randna()** function.

```{r}
se_randna <- sev::add_randna(se_filtered)

se_imp <- sev::impute_DEP(se_randna, randna = rowData(se_randna)$randna, fun = "mixed", mar = "knn", mnar = "QRILC")
```

### Possibility 3: Use Perseus method
A common approach is to replace NAs with small values, as missing data appears with a higher rate for low abundance proteins. The standard function in Perseus is, to replace missing values by drawing from a left shiftet gaussian distribution. 

```{r impute-perseus}
se_imp <- sev::impute_perseus(se_filtered)

```

## Expression test
Differential expression tests can be performed with the original *DEP* functions that use limma in the background.

```{r expression-test}

# Possibility 1: Use DEP method to test all contrasts
se_diff_all_contrasts <- test_diff(se_imp, type = "all")
se_diff_all_contrasts <- add_rejections(se_diff_all_contrasts, alpha = 0.05, lfc = 1)

# Possibility 2: Use DEP method to test all cases against a common control
se_diff_ctrl <- test_diff(se_imp, type = "control", control = "no_glom_heat")
se_diff_ctrl <- add_rejections(se_diff_ctrl, alpha = 0.05, lfc = 1)

# Possibility 3: Use DEP method to test specific contrast 
se_diff_spec <- test_diff(se_imp, type = "manual", test = "griesser_heat_vs_no_glom_heat")
se_diff_spec <- add_rejections(se_diff_spec, alpha = 0.05, lfc = 1)
```

## GO-term Enrichment
The *clusterProfiler::gseGO()* function is used to perform GO-term enrichment analysis. In theory, any numerical column can be selected as *col_name*, although only lfq_data or even better fold changes should be used. If no col_name is provided, all columns ending with "_diff" are selected. Results are stored in the *metadata* of the provided SummarizedExperiment. 

```{r GO-term-enrichment}

se_GO <- sev::se_GOE(se_diff_spec, col_names = "griesser_heat_vs_no_glom_heat_diff", OrgDb = "org.Hs.eg.db", pvalueCutoff = 0.5)

```

### Accessing GO-enrichment results
Enrichments are stored as lists within the *metadata* and can be accessed as such. To get the complete dataframe containing NES, p-value and similar information, the function *mcols()* can be used. 

```{r GO-plot}

#Access list of enrichment result:
metadata(se_GO)$GO_enrichment %>% names()

#Get enrichment result
GO_res <- metadata(se_GO)$GO_enrichment[[1]] %>% # select the first GO-term enrichment saved in the metadata. 
  mcols() %>% # get table of the results
  as.data.frame() # transform to dataframe

library(ggplot2)
# Plot
GO_res %>% filter(p.adjust < 0.05) %>% ggplot(aes(x = NES, y = Description, color = p.adjust))+
  geom_point()+
  theme_bw()+
  facet_wrap(. ~ ONTOLOGY, scales="free", ncol=1)

```

## Using iSEE to visualize results
SummarizedExperiments can be interactivey explored via *iSEE*. Additional containers are provided here, to plot volcanos, link genes to the NCBI database and to return a table of GO-term enrichments.

### Prepare data
Tp use the additional containers, it is necessary to register columns that store p-values and fold-changes, as well as to register the the metadata. The *se_to_isee()* function takes care of this. Be sure to select the right assay as the main assay before running the code (the imputation functions already do this for you).

```{r iSEE-prep}
# assay(se_GO) <- assays(se_GO)$imputed_perseus
isee_object <- sev::se_to_isee(se_GO)

```

### Create app with custom plots
You can start the app by calling or even store it as a Rda file for future use. 

```{r iSEE-app}
# create app with custom plots
app <- sev::isee_mini(isee_object)

```

## Minimize experiment
The input file might contain a lot of additional information that is not of interest for you. The *minimal_data()* function returns the object with only the bare necesseties. It is fairly easy (it only uses *dplyr::select()* internally), so have a look at it if you need to delete further columns (e.g. if you are only interested in certain comparisons).

```{r reduce-rowData}
minimal_data <- sev::make_mini(isee_object)
```

## Save results
SummarizedExperiments can be saved like any other R object and read back in via *load()*. Make sure to do this at the end of your analysis. 
```{r save-results}
save(isee_object, file = "../data/isee_object.rda")

load("../data/isee_object.rda")
```

## Session information
Include session info
```{r}
sessionInfo()
```

